Introducing the \NAME Framework\\ for Efficient Intrusion Detection\\ in the Internet of Things

### 2014-10-15 paper.tex
added previous evaluation

network usages: number of packets, number of bytes

processing time

image size

TODO: memory usage ?

2 algorithms \cite{ganeriwal2008reputation}

TODO: add 3rd ? \cite{krontiris2009cooperative}

>>>>

To evaluate the prototype we chose to use a system based on the Atmel
ATMEGA1284p micro-controller \cite{datasheet:atmega1284p} and the Digi XBee S2
ZigBee module \cite{manual:xbee}. To drive the hardware a minimalistic library
was used, wrapping the technical calls to the components in more easy to use
functions.

The reason for this at first sight unusual setup is simple: relying on basic
hardware and software allows us to validate that the generator is capable of
generating code even for the most basic environment available. Any step up,
both in hardware or software, would offer better and higher abstractions that
would make it easier to generate code for that situation. It shows that the
requirements of the generator towards the platform are minimal, don't rely on
advanced frameworks or operating systems and can therefore be applied in any
environment, targeting any platform.

For the evaluation we constructed a small WSN consisting of three nodes: an
end-device, a router and a coordinator. The end-device and the router were
completely autonomous nodes, while the coordinator consisted of an XBee module,
directly connected to a computer via a USB break-out board, allowing serial
access via a terminal.

We started from a basic application that measures light intensity and reports
it to the coordinator. The application was written both manually and generated.
On top of this baseline we added two intrusion detection related algorithms: a
heartbeat, that allows nodes to validate each other's continuous presence, and
a reputation-building algorithm that checks if a parent node is cooperative and
actually forwards messages with a destination further down the
network\cite{ganeriwal2008reputation}.

Three criteria were evaluated: the image size of the resulting compiled code,
the network usage in number of frames and bytes and the time required to
perform once cycle of the event loop. These metrics were collected in four
situations: without ID algorithms, with a heartbeat, with reputation-tracking
and with both algorithms implemented.

In case of the manual implementation, both algorithms were constructed as
standalone modules and sequentially called from the base-application's event
loop. The code generator was provided with FOO-lang descriptions of the
algorithms and generated all four cases.

Before reviewing the results, it is very important to put these results in a
proper perspective. Comparing manually written code to generated code is no
exact science. One can always argue that the quality of the two code bases is
not comparable and that the manual code always can be improved.

Still it might be interesting to look at the numbers, given that both code
bases are constructed with the same intentions and practices. We believe we
have achieved this and have taken honest decisions while constructing the
implementations, making them comparable. Numerical analysis of the
implementation offers us a preliminary estimation of the effect of our proposed
solution.

Tables \ref{tbl:manual} and \ref{tbl:generated} show the collected data
respectively for the manual and the generated implementation. The base case is
presented in absolute values, while the implementations with the added
algorithms are presented as relative values.

\begin{table}[H]
  \centering
  \begin{tabular}{lrrrr}
  \hline
      & base & heartbeat & reputation & both\\
  \hline
  size (bytes) & 10500 & 148\% & 127\% & 175\%\\
  frames & 20 & 255\% & 160\% & 315\%\\
  bytes & 476 & 406\% & 181\% & 487\%\\
  time ($\mu$s) & 48 & 196\% & 183\% & 310\%\\
  \hline
  \end{tabular}
  \caption{Results for the manual implementation.}
  \label{tbl:manual}
\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{lrrrr}
  \hline
         & base & heartbeat & reputation & both\\
  \hline
  size (bytes) & 10496 & 175\% & 156\% & 200\%\\
  frames & 20 & 245\% & 160\% & 275\%\\
  bytes & 476 & 399\% & 186\% & 454\%\\
  time ($\mu$s) & 48 & 252\% & 252\% & 288\%\\
  \hline
  \end{tabular}
  \caption{Results for the generated implementation.}
  \label{tbl:generated}
\end{table}

From these raw results, we learn that adding algorithms manually to the
base-application, results in a cumulative increase. The impact of both
algorithms is simply the sum of the impact of each algorithm by itself. In the
case of the time required to execute one cycle of the event loop, the total
time is even a bit higher than simply the sum.

In the case of the generated implementation, this no longer holds: although
that even the individual increases due to adding a single algorithm are higher
than in the manual case, the result for the implementation of both algorithms
is less than the sum of both. Here we clearly see the effect of reusing the
framework that comes with FOO-lang. The same goes for all other metrics. Maybe
most remarkable is the effect on the time of one event loop cycle: both
algorithms by itself add 150\% with respect to the base case, but when
combined, the impact hardly increases. Here we learn that the introduced
framework is responsible for the initial overhead, but it clearly pays for
itself when adding more algorithms.

Table \ref{tbl:summary} compares the two situations by subtracting the manual
case from the generated case, showing the impact of the code generation. A
relative value for the entire implementation with both algorithms is also
presented.

\begin{table}[H]
  \centering
  \begin{tabular}{lrrrrr}
  \hline
                & base & heartbeat & reputation & both  & result \\
  \hline
  size (bytes)  & -4    & 2822     & 3070       & 2664  & 115\%  \\
  frames        & 0     & -2       & 0          & -8    & 87\%   \\
  bytes         & 0     & -36      & 24         & -156  & 93\%   \\
  time ($\mu$s) & 0     & 27       & 33         & -11   & 93\%   \\
  \hline
  \end{tabular}
  \caption{Comparison of both implementations.}
  \label{tbl:summary}
\end{table}

When comparing the results of both implementations we first can conclude that
the generator produces exactly the same code for the base case. We couldn't
measure any real differences between both implementations.

Secondly, we see that the framework that comes with the generated code adds to
the size of the resulting image. But the cost is almost constant and is
relatively lower when algorithms are combined. With roughly 3KB of additional
overhead, or 15\% in this simple case, this impact is affordable.

From a functional point of view, the generated code lives up to the
expectactions: both the number of frames as the bytes sent benefit from well
organized code and reuse of a common framework.

Finally we see that both algorithms by itself add an equal amount of time to
the processing of a single event loop cycle, but when combined, the event loop
is about 7\% faster compared to the manual case.

Without any effort to optimize the framework code, nor leveraging knowledge
about the specific algorithms, this very basic generated code shows that the
proposed principles are not only theoretically feasible but actually result in
interesting improvements.
